Two Problems for Sophistication

### 2015-05-08 Facticity/Facticity/paper.tex
changes...

% --- DELETE BEFORE SUBMISSIONS ---
% \pagestyle{headings}

@@

\begin{abstract}
Kolmogorov complexity measures the amount of information in data, but does not distinguish structure from noise. Kolmogorov's definition of the \emph{structure function} was the first attempt to measure only the structural information in data, by measuring the complexity of the smallest model allowing for optimal compression of the data. Since then, many variations of this idea have been proposed, for which we use \emph{sophistication} as an umbrella term. We describe two fundamental problems with all existing proposals, showing each  to be \emph{unsound}. Consequently, we put forward the view that the problem is fundamental: it may be impossible to objectively quantify the sophistication.
\end{abstract}

@@

The Kolmogorov complexity $C(x)$ of a binary string $x$, is (roughly) the length of the shortest computer program that will output $x$. This length depends on which programming language is used, but, as shown by a central result known as the Invariance Theorem \cite[Section~2.1]{li1993introduction}, the language affects the value only by a constant, independent of $x$. This means that for sufficiently complex objects, the choice of programming language becomes irrelevant and Kolmogorov complexity becomes an \emph{objective} measure of the complexity of binary strings, a property \emph{of the data.} A definition of sophistication $S(x)$ in the spirit of Kolmogorov complexity should have similar robustness guarantees:

\begin{enumerate}
\item $S(x)$ should count the bits required for an effective description of the structural properties of a binary string.
\item An analogue of invariance should hold: there must be strict limits on how much sophistication can be affected by a change in programming language.
\item There should be no constant $c$ such that $S(x)\le c$ for every input $x$. If sophistication is bounded, then knowing its value under one programming language provides no constraints on its value under another language (except that it's also bounded). Moreover, a bounded sophistication would be at odds with the intuition that there is no limit to the amount of structure that a binary string can exhibit.
\item Similarly, there should be no constant $c$ such that $C(x)-S(x)\le c$ for all $x$, because then sophistication would be equivalent to Kolmogorov complexity. 
\end{enumerate}
There have been many related proposals for such a measure, all based on a \emph{two-part code}: rather than encoding $x$ by means of a single computer program, we encode a \emph{model} for $x$ in the first part of the code, that is interpreted as a representation of $x$'s structural properties. The model does not fully specify $x$, but when combined with the second part of the code, which specifies the noise, the original string becomes fully determined. Thus, the model can be interpreted as a function mapping the noise to the original object.

@@

Let $\B = \{0,1\}^*$. We deal with partial computable functions $\phi: \B \to \B$, which we will also call \emph{models}. A function is called \emph{prefix} if its domain, $\dom{f} = \{y : f(y) \neq \infty\}$ is a prefix free set, i.e. no string in $\dom{f}$ is a prefix of another. A function $f$ is \emph{total} if $\dom{f} = \B$.

@@

A \emph{model class} is a set of model indices. We distinguish four important model classes: (1) The indices of the partial computable functions $\C=\N$. (2) The total functions $\T=\{i:\tn{$\phi_i$ is total}\}$. Note that $\T$ is not computably enumerable. (3) $\K$ is an enumerable set such that $\{\phi_i:i\in\K\}$ is the set of all partial computable prefix functions. (4) The finite sets: $\F$ is an enumerable set such that $\{\phi_i:i\in\F\}$ contains a uniform code for every finite set. A uniform code for a finite set $S$ is a total computable surjective function $f:\{0,1\}^{\lceil\log|S|\rceil}\to S$.

@@

For technical reasons, we deviate slightly from the traditional notation of Kolmogorov complexity: we define the $\M$-Kolmogorov complexity of a binary string $x$, with respect to a model class $\M$ and a numbering $\psi$, as follows $C^{\M,\psi}(x)=\min\{\bar\imath y:\psi_i(y)=x,i\in\M\}$. This means that $C^\C(x)$ corresponds to the plain Kolmogorov complexity, traditionally denoted $C(x)$, and $C^\K(x)$ corresponds to the prefix-free Kolmogorov complexity $K$. This notation also allows us to represent the smallest possible description of $x$ using a single model $\phi_i$ as $C^{\{i\}}(x)$. Note that this includes the description of $i$ as well as its shortest program for $x$.

Finally we not that the choice of numbering, combined with a choice of prefix function, serves the same purpose that is expressed in the choice of a universal Turing machine. Requiring an acceptable numbering and a computable prefix function is equivalent to choosing a universal Turing machine: this requirement suffices to make the Kolmogorov complexity invariant. As we shall see, however simply choosing a universal Turing machine is not enough for a proper treatment of sophistication: we must choose our numbering carefully. Thus, we phrase our requirement for an invariant sophistication with respect to the \emph{numbering} chosen.    There should be strict guarantees on how much the value of $S(x)$ can changes if the numbering is varied.

@@

The simplest approach to sophistication would simply be to `open up' the Kolmogorov complexity and to see which program on the universal Turing machine achieves the smallest description length: the program that \emph{witnesses} the Kolmogorov complexity. Since any UTM is defined in terms of a two-part coding, such programs necessarily consist of a model and an input. While this approach is not literally used anywhere in the literature, it is an illustrative point to start our investigation. As is common, we will allow all witnesses within a constant slack term $c$ from the Kolmogorov complexity, and choose the one with the smallest model.

\begin{definition}[Index sophistication]
\[
s_{\text{index}}^{\M_0,\M_1,\psi,c}(X) = \min\left\{ |i| : \i in \M_0, C^{\{i\}, \psi}(x) \leq C^{\M_1, \psi}(x)+c   \right\}
\]
\end{definition}

Koppel and Atlan's treatment \cite{koppelSoph1988,koppel1991almost}, where the name \emph{sophistication} originates, follows this basic logic, although it contains many idiosyncracies like the use ot monotonic models, and an inherent extension to infinite strings. As the subsequent history of sophistication has discarded these, we will not discuss them here.

In \cite{antunes2009sophistication} Koppel's principle is first recast to finite strings. The definition used is similar to $s_\text{index}^{\T, \C,\psi,c}$, with one exception: the total complexity of a witness $(i,y)$ is measured as $|i|+|y|$ without the cost of delimiting the two. This differences are not relevant to the point we make in this section. The restriction to total classes is a common technique to avoid underfitting, which we will discuss in the next section.

@@

Note that such perverse numberings are not an issue for Kolmogorov complexity. If our universal Turing machine uses an inefficient numbering, we can switch to a moe efficient one at only a constant cost. This gives us an insight into what happens inside Kolmogorov complexity. We may define the program $\bi y$ using a model $i$ and an input $y$, but if there's any inefficiency in the description of $i$, the shortest program will be achieved by using $i$ to define a more efficient model encoding, and encoding both the model and its input in $y$. For the Kolmogorov complexity, such considerations are academic, since we only look at the total length of the program. For sophistication, however, they are crucial.

@@

Second, we can stick to using the length of the index as the measure of sophistication, but restrict the allowed numberings. This approach is taken is in \cite{adriaans2012facticity}. Adriaans  defines \emph{facticity} as $s_\text{index}^{\C,\C,\psi,0}$, but only allows numberings that are \emph{faithful}: for all $i$ there is a $j$ such that $\psi_i=\psi_j$ and $|j|\le C^{\C}(\psi_j)+c$, for some constant $c$.

@@

Because of these issues, rather than index sophistication we will use the following definition of sophistication in the remainder of this article. It is parameterised by a model class, a numbering, and a slack parameter $c$ that determines the maximal allowed slack between the two-part code length using a representation and the Kolmogorov complexity.
\[
S^{\M,\psi,c}(x)=\min\left\{C^\K(\phi_i):i\in\M,C^{\{i\},\psi}(x)\le C^{\M,\psi}(x)+c\right\}.
\]

@@

In the previous section, we saw the first glimpse of the delicate balance between the two code components in sophistication. We will investigate this balance, starting with the variant $S^{\K,\psi,c}$. For reasons which will become clear, this variant is not used anywhere in the literature, but it helps to illustrate the issues we wich to discuss.

@@

\begin{restatable}[Underfitting]{lemma}{underfitting}
Let $\M$ be a model class containing a universal model $\phi_u$, with the property that $\exists c \forall i \in \M : C^{\{u\}}(x) \leq C^{\{i\}}(x) + c \p$. Then, there exist numberings such that $\s^\M$ is bounded.
\end{restatable}

@@

The proof of both these lemmas relies on a simple principle: for certain subclasses $\M' \subset \M$, there exist numberings which, when used, have the effect of penalizing the codelength $C^\K(\phi_i)$ for any model outside $\M'$ by an arbitrary constant amount. We can use this penatly to effectively `push' these models outside of the range of considered models, ensuring that, under this numbering, a model in $\M'$ always determines the sophistication.

For the lemmas above this means setting $\M' = {u}$, with $\phi_u$ a universal model, or letting $\M'$ be the set of all singletons. The requirements for $\M'$ are strict, and slightly complex. The following lemma illustrates the principle.

@@

Now, by Lemma~\ref{lemma:thecoolone} there is an numbering $\psi_1,\psi_2,\ldots$ such that $\s^{\M,\phi}(x)=|\bar 0|+S^{\M',\psi}(x)$. We observed that $|C^{\K}(f)-C^{\K}(x)|\le c$ for any $f\in\M'$, so $\s^{\M',\psi}(x)\ge C^{\K}(x)-c$. This proves the lemma.
\end{proof}

\noindent Thus, in this balanced sophistication, there is no invariance: all information can be seen as structure, or as noise, depending on the numbering. To avoid these issues, existing proposals upset the balance to exclude or penalize at least the singleton models and the universal models.

@@

We will now review the treatments in the literature that show overfitting. The first is the structure function, proposed by Kolmogorov himself. Most likely the first attempt at separating structure from noise in an object manner. Kolmogorov defined the function
\[
h_x(\alpha) = \min \left \{\log |S| : S \ni x, C^\K(s) \leq \alpha \right \} 
\]
and suggested that the smallest set for which $K(S) + \log|S| \leq C^\K(S) + c$ for some pre-chosen constant$c$ can be seen as capturing all the structural information in $x$. \footnote{\pb{Discuss how much of this Kolmogorov believed himself and note that this was unpublished work.}} This is equivalent to the sophistication $\s^{\F,\K,\psi,c}$. Lemma~\ref{lemma:underfitting} shows us that given $c$, there are numberings for which this sophistication is always equal, up to a constant, to $C^\K(x)$. Thus, while the general shape of the structure function might tell an interesting story, its sophistication function does not satisfy our requirements.

$\s^{\F,\K,\psi,c}$ is also used in \cite{antunes2013sophistication}. \pb{This version doesn't work at all\ldots}

Also the regular and naive versions from \cite{mota2013sophistication} \ldots

the structure function \cite{cover1985kolmogorov}

the sophistication as defined in \cite{mota2013sophistication} (and the naive soph?)

the minimal sufficient algorithmic statistic \cite{gacs2001algorithmic}

Vereshchagin's strongly sufficient statistic \cite{vereshchagin2013algorithmic}. \pb{fix citation}

@@

It may be argued that the slack parameter $c$ in the sophistication, that determines the allowed gap between a candidate representation and the complexity, should be allowed to depend on the numbering, but this dependence has not been mentioned in the literature and there is no clear way to determine how this constant should be determined on the basis of a given numbering.

In traditional statistics, overfitting is often addressed by imposing a penalty on complex models. As we have seen, a strong penalty, such as the one imposed by an inefficient prefix encoding of the model, will immediately cause underfitting. A more subtle approach is to use $\C$ as a model class rather than $\K$, thus allowing representations that are not self-delimiting. The gap between self-delimiting and non self-delimiting descriptions grows without bound 
\cite[Section~4.5.5]{vitanyi2004meaningful}, so that for sufficiently complex inputs all information will have to be placed in the noise part of the code, eliminating the singletons as viable candidates. This approach is taken by Vit\'anyi \cite{vitanyi2004meaningful} and by Adriaans \cite{adriaans2012facticity}. Such measures thus help reduce the overfitting problem, but they only increase the tendency to underfit. We also pay the price that the models can no longer be equated with probability measures, removing the link to traditional statistics and probability theory.

\subsection{Underfitting}

As mentioned in the related work, underfitting is a widely acknowledged problem for sophistication, and most proposals avoid it by limiting the class of allowed models, to exclude universal models. The following theorem shows that this does indeed yield an unbounded sophistication.

\begin{restatable}{theorem}{dogfood}
There are infinitely many $x$ with $\s^{\F,\phi,c}(x) \geq C^{\K}(x)/5$, and also infinitely many $x$ with $\s^{\T,\phi,c}(x) \geq C^{\K}(x)/6$.
\end{restatable} 
\noindent Thus, there are no longer any numberings for which sophistication is bounded. However, as we show below, the resulting sophistication is not an intuitive measure of the amount of structure in the data. Moreover, the balance is pushed back in the direction of overfitting, which is only exacerbated by this change. Only one model class eliminates both the singletons and the universal model: $\T$. The only proposal that uses an efficient model representation \emph{and} excludes the universal models \emph{and} excludes the singletons is \cite{vitanyi2004meaningful}. While this invalidates straightforward proofs of boundedness, there is no suggestion that $S^{\T, \phi, c}(x)$ is actually invariant.

@@

\begin{definition}[Depth\cite{bennett1988logical,antunes2006computational}]\belowdisplayskip=-12pt
Let $U$ be some universal Turing machine, so that $U(\bar\imath y) = \phi_i(y)$. Let $U^t$ be a simulation of this machine, which is allowed to run for at most $t$ steps, and returns $0$ if it has not yet finished at that point. Let $C^\M_t(x) = \min\{|\bar\imath y| : U^t(\bar\imath y) = x, \phi_i \in \M\}$. The \emph{$c$-depth} is $d^{\M,c}(x) = \min \left\{t : C^\M_t(x) - C^\M(x) \leq c \right\}$.
\end{definition}
Deep strings are those that can only be optimally compressed with a great investment of time. We note that it is exceedingly unlikely that a deep string is sampled from a shallow distribution \cite{bloem2014safe,bennett1988logical}. 
\cite{antunes2013sophistication} also investigates the relation between $S(x)$ and $d(x)$.

@@

This clearly contradicts the motivating principle of sophistication, which is to capture the number of bits required to express \emph{all} regularity in the data, not just the regularity that is difficult to describe. 
In this setting, a string high-sophistication not only has a lot of structure, but it also has the property that it can be compressed far better with partial functions than with total. That is, it is non-typical for the model class $\T$. This suggests that the `non-stochastic' property of strings with high sophistication \cite{shen1983concept,vereshchagin2004kolmogorov} says more about depth and totality than it does about structure and noise.

@@

This idea is bold, but not necessarily unreasonable. Nevertheless, in this section we offer the opinion that such a clean-cut separation \emph{cannot} be made to work. We provide two arguments to support this belief.

For the first argument, we take a generative perspective. We can generate data from a model in $\phi_i, i \in \K$ by feeding it random bits until it produces an output. We will call the resulting probability distribution $p_i(x) = \sum_{y:\phi_i(y) = x} 2^{-|y|}$. In this scenario, a sophistication could be called \emph{consistent} if, for data of sufficiently high complexity, it accurately reflects the complexity of that distribution from which it originated. Suppose that the data is sampled from the universal distribution: let $\phi_u(\bar\imath y)=\phi_i(y)$ and sample from $p_u$ as described above. Then the initial random bits will determine the prefix encoded index $\bar\imath$ of the function $\phi_i$ that $\phi_u$ will subsequently emulate, and the remaining bits are used as inputs to $\phi_i$. We now ask, what should be the sophistication of the resulting data?

@@

The second reason we believe sophistication will elude definition is more technical. Given an input $x$ and some model class, say $\C$, consider the set of all possible two-part representations of $x$. Now, when the numbering of partial computable functions is changed, the codelength of the model part of all these representations will change. This is illustrated in the right graph in Figure~\ref{fig:diagram}. However, the Invariance Theorem expresses that this change is limited by a constant term that does not depend on $x$. So for sufficiently large $x$, the horizontal shift of all the representations will be relatively small. However, even this small shift can push some representations out of the acceptable region (indicated by the dashed line), and pull others in. This may lead to a different representation determining the sophistication, one whose \emph{total} codelength is close to what it was before, but whose \emph{model} codelength can be anywhere between $0$ and $C^\C(x)$. If such jumps can occur, the sophistication is not invariant. And while we cannot \emph{prove} in general that such jumps can always occur, there seems to be no reason to believe that they don't.

For this reason, we take a skeptical view of sophistication. Note that part of the theory is fine: there is nothing wrong with evaluating models for the data by comparing their two-part code lengths. In fact, the randomness deficiency $\delta(\phi_i,x)=-\log p_i(x)-C^\K(x\mid i)$ of a model $\phi_i$ has a direct statistical interpretation as a measure of counterevidence against the hypothesis that it generated the data---if $p_i$ truly generated the data, then the probability that the randomness deficiency is more than $k$ bits is less than $2^{-k}$ \cite[Lemma~6]{bloem2014safe}. In the Monet example above, this will allow us to disqualify the model expressing that the data is actually, say, a recording of Jazz music.

>>>>

@@

\begin{abstract}
Kolmogorov complexity measures the amount of information in data, but does not distinguish structure from noise. Kolmogorov's definition of the \emph{structure function} was the first attempt to measure only the structural information in data, by measuring the complexity of the smallest model allowing for optimal compression of the data. Since then, many variations of this idea have been proposed, for which we use \emph{sophistication} as an umbrella term. We describe two fundamental problems with existing proposals, showing many of them to be unsound. Consequently, we put forward the view that the problem is fundamental: it may be impossible to objectively quantify the sophistication.
\end{abstract}

@@

The Kolmogorov complexity $C(x)$ of a binary string $x$, is (roughly) the length of the shortest computer program that will output $x$. This length depends on which programming language is used, but, as shown by a central result known as the invariance theorem \cite[Section~2.1]{li1993introduction}, the language affects the value only by a constant, independent of $x$. This means that for sufficiently complex objects, the choice of programming language becomes irrelevant and Kolmogorov complexity becomes an \emph{objective} measure of the complexity of binary strings, a property \emph{of the data.} A definition of sophistication $S(x)$ in the spirit of Kolmogorov complexity should have similar robustness guarantees:

\begin{enumerate}
\item $S(x)$ should count the bits required for an effective description of the structural properties of a binary string.
\item An analogue of invariance should hold: there must be strict limits on how much sophistication can be affected by a change in programming language.
\item There should be no constant $c$ such that $S(x)\le c$ for every input $x$. If sophistication is bounded, then knowing its value under one programming language provides no constraints on its value under another language (except that it's also bounded). Moreover, a bounded sophistication would be at odds with the intuition that there is no limit to the amount of structure that a binary string can exhibit.
\item Similarly, there should be no constant $c$ such that $C(x)-S(x)\le c$ for all $x$, because then sophistication would be equivalent to Kolmogorov complexity. 
\end{enumerate}
There have been many related proposals for such a measure, all based on a \emph{two-part code}: rather than encoding $x$ by means of a single computer program, we encode a \emph{model} for $x$ in the first part of the code, which is interpreted as a representation of $x$'s structural properties. The model does not fully specify $x$, but when combined with the second part of the code, which specifies the noise, the original string becomes fully determined. Thus, the model can be interpreted as a function mapping the noise to the original object.

@@

Let $\B = \{0,1\}^*$. We deal with partial computable functions $f: \B \times \B \to \B$, which we will also call \emph{models}. A function is called \emph{prefix} if its domain, $\dom{f} = \{y : f(y; \cdot) \neq \infty\}$ is a prefix free set, i.e. no string in $\dom{f}$ is a prefix of another. A function $f$ is \emph{total} if $\dom{f} = \B$. In most cases, we do not use the second argument, and we adopt the convention that $f(x) = f(x, \epsilon)$.

@@

A \emph{model class} is a set of model indices. We distinguish four important model classes:
\begin{itemize} 
  \item The indices of the partial computable functions $\C=\N$.
  \item The total functions $\T=\{i:\tn{$\phi_i$ is total}\}$. Note that $\T$ is not computably enumerable.
  \item $\K$ is an enumerable set such that $\{\phi_i:i\in\K\}$ is the set of all partial computable prefix functions.
  \item The finite sets: $\F$ is an enumerable set such that $\{\phi_i:i\in\F\}$ contains a uniform code for every finite set. A uniform code for a finite set $S$ is a total computable surjective function $f:\{0,1\}^{\lceil\log|S|\rceil}\to S$.
\end{itemize}

@@

For technical reasons, we deviate slightly from the traditional notation of Kolmogorov complexity: we define the $\M$-Kolmogorov complexity of a binary string $x$, with respect to a model class $\M$ and a numbering $\psi$, as follows $C^{\M,\psi}(x\mid z)=\min\{\bar\imath y:\psi_i(y; z)=x,i\in\M\}$, with $C^\M(x) = C^\M(x\mid \epsilon)$. This means that $C^\C(x)$ corresponds to the plain Kolmogorov complexity, traditionally denoted $C(x)$, and $C^\K(x)$ corresponds to the prefix-free Kolmogorov complexity $K$. This notation also allows us to represent the smallest possible description of $x$ using a single model $\phi_i$ as $C^{\{i\}}(x)$. Note that this includes the description of $i$ as well as its shortest program for $x$.

Finally we note that the choice of numbering, combined with a choice of prefix function, serves the same purpose that is normally expressed in the choice of a Turing machine. Requiring an acceptable numbering and a computable prefix function is equivalent to choosing a universal Turing machine: this requirement suffices to make the Kolmogorov complexity invariant. As we shall see, however, this is not enough for a proper treatment of sophistication: we must choose our numbering with care. Thus, we phrase our requirement for an invariant sophistication with respect to the \emph{numbering} chosen. There should be strict guarantees on how much the value of $S(x)$ can change if the numbering is varied.

@@

The simplest approach to sophistication would simply be to `open up' the Kolmogorov complexity and to see which program on the universal Turing machine achieves the smallest description length: the program that \emph{witnesses} the Kolmogorov complexity. Since any UTM is defined in terms of a two-part coding, such programs necessarily consist of a model and an input. As is common, we will allow all witnesses within a constant slack term $c$ from the Kolmogorov complexity, and choose the one with the smallest model.

\begin{definition}[Index sophistication]
\[
\s_{\text{index}}^{\M_1,\M_2,\psi,c}(X) = \min\left\{ |i| : \i \in \M_1, C^{\{i\}, \psi}(x) \leq C^{\M_2, \psi}(x)+c   \right\}
\]
When $\M_1 = \M_2$, we will abbreviate with $\s^{\M_1,\psi,c}$
\end{definition}

Koppel and Atlan's treatment \cite{koppelSoph1988,koppel1991almost}, where the name \emph{sophistication} originates, follows this basic logic, although it contains many idiosyncracies like the use of monotonic models, and an inherent extension to infinite strings. As the subsequent history of sophistication has discarded these, we will not discuss them here.

In \cite{antunes2009sophistication,antunes2013sophistication} Koppel's principle is recast to finite strings. The definition used is similar to $s_\text{index}^{\T, \C,\psi,c}$, with one exception: the total complexity of a witness $(i,y)$ is measured as $|i|+|y|$ without the cost of delimiting the two. This difference is not relevant to the point we make in this section. The restriction to total classes is a common technique to avoid underfitting, which we will discuss in the next section.

@@

Note that such perversely inefficient numberings are not an issue for Kolmogorov complexity. If our universal Turing machine uses an inefficient numbering, we can switch to a more efficient one at only a constant cost. This gives us an insight into what happens inside Kolmogorov complexity. We may define the program $\bi y$ using a model $i$ and an input $y$, but if there's any inefficiency in the description of $i$, the shortest program will be achieved by using $i$ to define a more efficient model encoding, and encoding both the model and its input in $y$. For the Kolmogorov complexity, such considerations are academic, since we only look at the total length of the program. For sophistication, however, they are crucial.

@@

Second, we can stick to using the length of the index as the measure of sophistication, but restrict the allowed numberings. This approach is taken is in \cite{adriaans2012facticity}. Adriaans  defines \emph{facticity} as $\s_\text{index}^{\C,\psi,0}$, but only allows numberings that are \emph{faithful}: for all $i$ there is a $j$ such that $\psi_i=\psi_j$ and $|j|\le C^{\C}(\psi_j)+c$, for some constant $c$.

@@

Because of these issues, rather than index sophistication we will use the following definition of sophistication in the remainder of this article. It is parameterised by a model class, a numbering, and a slack parameter $c$ that determines the maximal allowed slack between the two-part code length using a representation and the Kolmogorov complexity.
\[
\s^{\M_1,\M_2,\psi,c}(x)=\min\left\{C^\K(\phi_i):i\in\M_1,C^{\{i\},\psi}(x)\le C^{\M_2,\psi}(x)+c\right\}.
\]

@@

In the previous section, we saw the first glimpse of the delicate balance between the two code components in sophistication. We will investigate this balance, starting with the variant $\s^{\K,\psi,c}$. For reasons which will become clear, this variant is not used anywhere in the literature, but it helps to illustrate the issues we wish to discuss.

@@

\begin{restatable}[Underfitting]{lemma}{underfitting}
Let $\M$ be a model class containing a universal model $\phi_u$, with the property that $\exists c \forall i \in \M : C^{\{u\}}(x) \leq C^{\{i\}}(x) + c \p$. Then, there exist numberings such that $\s^\M$ is bounded.\label{lemma:underfitting}
\end{restatable}

@@

The proof of both these lemmas relies on a simple principle: for certain subclasses $\M' \subset \M$, there exist numberings which, when used, have the effect of penalizing the codelength $C^\K(\phi_i)$ for any model outside $\M'$ by an arbitrary constant amount. We can use this penalty to effectively `push' these models outside of the range of considered models, ensuring that, under this numbering, a model in $\M'$ always determines the sophistication.

For the lemmas above, this means setting $\M' = {u}$, with $\phi_u$ a universal model, or letting $\M'$ be the set of all singletons. The requirements for $\M'$ are strict, and slightly complex. The following lemma gives a set of sufficient conditions.

@@

Now, by Lemma~\ref{lemma:thecoolone} there is a numbering $\psi_1,\psi_2,\ldots$ such that $\s^{\M,\phi}(x)=|\bar 0|+S^{\M',\psi}(x)$. We observed that $|C^{\K}(f)-C^{\K}(x)|\le c$ for any $f\in\M'$, so $\s^{\M',\psi}(x)\ge C^{\K}(x)-c$. This proves the lemma.
\end{proof}

\noindent Thus, in this balanced sophistication, there is no invariance: all information can be seen as structure, or as noise, depending on the numbering. To avoid these issues, existing proposals upset the balance to exclude or penalize the universal models, and possibly the singleton models.

@@

We will now review the treatments in the literature that show overfitting. The first is the structure function, proposed by Kolmogorov: most likely the first attempt at separating structure from noise in an objective manner. Kolmogorov defined the function
\[
h_x(\alpha) = \min \left \{\log |S| : S \ni x, C^\K(s) \leq \alpha \right \} 
\]
and suggested that the smallest set for which $K(S) + \log|S| \leq C^\K(S) + c$ for some pre-chosen constant $c$ can be seen as capturing all the structural information in $x$ \cite{cover1985kolmogorov}. This is equivalent to the sophistication $\s^{\F,\K,\psi,c}$. Lemma~\ref{lemma:underfitting} shows us that given $c$, there are numberings for which this sophistication is always equal, up to a constant, to $C^\K(x)$. Thus, while the general shape of the structure function might tell an interesting story, its sophistication function is either not invariant, or vacuous.

\pb{Discuss the minimal sufficient algorithmic statistic \cite{gacs2001algorithmic}}

@@

It may be argued that the slack parameter $c$ in the sophistication, that determines the allowed gap between a candidate representation and the complexity, should be allowed to depend on the numbering, but this dependence has not been mentioned in the literature and there is no clear way to determine how this constant should be chosen on the basis of a given numbering.

In traditional statistics, overfitting is often addressed by imposing a penalty on complex models. As we have seen, a strong penalty, such as the one imposed by an inefficient prefix encoding of the model, will immediately cause underfitting. A more subtle approach is to use $\C$ as a model class rather than $\K$, thus allowing representations that are not self-delimiting. The gap between self-delimiting and non self-delimiting descriptions grows without bound 
\cite[Section~4.5.5]{li1993introduction}, so that for sufficiently complex inputs, some information will have to be placed in the noise part of the code, eliminating the singletons as viable candidates. This approach is taken by Vit\'anyi \cite{vitanyi2004meaningful} and by Adriaans \cite{adriaans2012facticity}. Such measures thus help reduce the overfitting problem, but they only increase the tendency to underfit. We also pay the price that the models can no longer be equated with probability measures, removing the link to traditional statistics and probability theory.

\subsection{Underfitting}
\label{section:underfitting}

Underfitting is a widely acknowledged problem for sophistication, and most proposals avoid it by limiting the class of allowed models to exclude universal models. The following theorem shows that this does indeed yield an unbounded sophistication.

\pb{Recheck what was known already}.
\begin{restatable}{theorem}{dogfood}
There are infinitely many $x$ with $\s^{\F,\phi,c}(x) \geq C^{\K}(x)/5$, and also infinitely many $x$ with $\s^{\T,\phi,c}(x) \geq C^{\K}(x)/6$.
\end{restatable} 
\noindent Thus, there are no longer any numberings for which sophistication is bounded. However, as we show below, the resulting sophistication is not an intuitive measure of the amount of structure in the data. Moreover, the balance is pushed back in the direction of overfitting, which is only exacerbated by this change. Only one model class eliminates both the singletons and the universal model: $\T$. The only proposal that uses an efficient model representation \emph{and} excludes the universal models \emph{and} excludes the singletons is \cite{vitanyi2004meaningful}. While this invalidates straightforward proofs of boundedness, there is no suggestion that $S^{\T, \phi, c}(x)$ is actually invariant.

@@

\begin{definition}[Depth\cite{bennett1988logical,antunes2006computational}]\belowdisplayskip=-12pt
Let $U$ be some universal Turing machine, so that $U(\bar\imath y) = \phi_i(y)$. Let $U^t$ be a simulation of this machine, which is allowed to run for at most $t$ steps, and returns $0$ if it has not yet finished at that point. Let $C^\M_t(x) = \min\{|\bar\imath y| : U^t(\bar\imath y) = x, \phi_i \in \M\}$. The \emph{$c$-depth} is $d^{\M,c}(x) = \min \left\{t : C^\M_t(x) - C^\M(x) \leq c \right\}$.
\end{definition}
Deep strings are those that can only be optimally compressed with a great investment of time. We note that it is exceedingly unlikely that a deep string is sampled from a shallow distribution \cite{bloem2014safe,bennett1988logical}.

@@

The relation between $S(x)$ and $d(x)$ is also investigated in \cite{antunes2013sophistication}, where it is shown that within a logarithmic error term on the sophistication and the slack, they are identical. Our point here is not that the two functions are similar, but that for all practical strings, the sophistication is bounded.This contradicts the intuition that sophistication measures structure, as it seems to suggest that all strings we can possibly hope to encounter contain no structure, save a constant amount. The alternative is that under other numberings these strings \emph{would} have structure, but then the sophistication would no longer be invariant.

As for the strings with high sophistication, they have the property that they can be compressed far better with partial functions than with total: they are non-typical for the model class $\T$. This suggests that the `non-stochastic' property of strings with high sophistication \cite{shen1983concept,vereshchagin2004kolmogorov} says more about depth and totality than it does about structure and noise.

\pb{Vereshchagin's strongly sufficient statistic \cite{vereshchagin2013algorithmic}}

\subsection{Other variants}

By moving away from the idea of two-part coding, the mechanics of lemma~\ref{lemmathecoolone} can be avoided. In \cite{mota2013sophistication}, the \emph{naive sophistication} is introduced:
\[
\s_\text{naive}^{\psi, c}(x) = \min\left\{C^\K(x) : \log |S| - C^\K(x|s) \leq c \right\} \p
\] 
The condition is that the \emph{randomness deficiency} of $S$, the extent to which $x$ is typical for $S$, is constant. This is equivalent to the algorithmic minimal sufficient statistic based on the $\beta_x$ function from \cite{vereshchagin2004kolmogorov}. While the logic of Lemma~\ref{lemma:overfitting} does not work for this form, it is still an open question if it is invariant.

Another approach is the \emph{coarse sophistication} \cite{antunes2009sophistication}. The simplest definition comes from \cite{mota2013sophistication}:
\[
\s_\text{coarse}(x) = \min_c\left\{\s^{\F,\K,\psi,c}(x) + c \right\} \p
\]
Again, this variant avoids the pitfalls of Lemma~\ref{lemma:overfitting}. If a set of $x$s has witnesses $S_x$ that are as good as the singleton set, up to a constant, but with smaller $K(S)$ by more than a constant, the added penalty $c$ will eventually be much less than the gain for the simpler witness, and the singletons will not determine the coarse sophistication. However, whether this form is invariant is again an open question. We also note that when extended from $\F$ to $\T$, it runs into the same problem as discussed in Section~\ref{section:underfitting}

Finally we note that coarse sophistication moves further away from the original philosophy of sophistication than may at first be apparent\ldots \pb{Show, with some rewriting, how far this really deviates form the original sophistication.}

Finally, we must mention \emph{effective complexity} \cite{gellmann1996information}. This measure is unique, in that it was formulated from the perspective of physics. Nevertheless, it fits the mold of sophistication very neatly. The models used are computable probability distributions on finite sets, and all candidates that come within a constant of the Kolmogorov complexity are considered. The complexity of the model is correctly measured by its Kolmogorov complexity. We note that this means that Lemmas \ref{lemma:underfitting} and \ref{lemma:overfitting} both apply. Unlike other sophistication measures, here it is not the smallest model which is chosen, but the one which reproduces the data with in the shortest time. Thus, if there are multiple candidates, this approach would likely favor the singleton sets. In \cite{gell2004nonextensive}, the authors abandon this approach, and note the choice from the set of candidates is a subjective one, which depends on context.

@@

This idea is bold, but not necessarily unreasonable. Nevertheless, in this section we offer the opinion that such a clean-cut separation \emph{cannot} be made to work. We provide three arguments to support this belief.

For the first argument, we take a generative perspective. We can generate data from a model in $\phi_i, i \in \K$, by feeding it random bits until it produces an output. We will call the resulting probability distribution $p_i(x) = \sum_{y:\phi_i(y) = x} 2^{-|y|}$. In this scenario, a sophistication could be called \emph{consistent} if, for data of sufficiently high Kolmogorov complexity, it accurately reflects the complexity of that distribution from which it originated. Suppose that the data is sampled from the universal distribution: let $\phi_u(\bar\imath y)=\phi_i(y)$ and sample from $p_u$ as described above. Then the initial random bits will determine the prefix encoded index $\bar\imath$ of the function $\phi_i$ that $\phi_u$ will subsequently emulate, and the remaining bits are used as inputs to $\phi_i$. We now ask, what should be the sophistication of the resulting data?

@@

The second reason we believe sophistication will elude definition is more technical. Given an input $x$ and some model class, say $\C$, consider the set of all possible two-part representations of $x$. Now, when the numbering is changed, the codelength of the model part of all these representations will change. This is illustrated in the right graph in Figure~\ref{fig:diagram}. The invariance theorem expresses that this change is limited by a constant term that does not depend on $x$. So for sufficiently large $x$, the horizontal shift of all the representations will be relatively small. However, even this small shift can push some representations out of the acceptable region (indicated by the dashed line), and pull others in. This may lead to a different representation determining the sophistication, one whose \emph{total} codelength is close to what it was before, but whose \emph{model} codelength can be anywhere between $0$ and $C^\C(x)$. If such jumps can occur, the sophistication is not invariant. And while we cannot \emph{prove} in general that such jumps can always occur, there seems to be no reason to believe that they don't. Indeed, in \cite{antunes2013sophistication} it is shown that logarithmic changes in the slack parameter can already cause these effects.
